---
title: "COVID19 projections using Generalized Richards Model"
description: "Making COVID19 forecasts on confirmed cases and deaths around the world using Generalized Richards Model (GRM)"
categories:
  - R
  - "COVID-19"
author:
  - name: Ankur Singhal 
    affiliation: Empirical Solutions Consulting, LLC
creative_commons: CC BY-SA
date: "2020-04-14"
base_url: https://aksinghal86.github.com/io/covid19
slug: aksinghal862020covid19GRM
twitter:
  creator: "@skepticalAnkur"
output:
  distill::distill_article:
    toc: true
    toc_depth: 2
    self_contained: false
repository_url: https://github.com/aksinghal86/covid19/blob/master/_posts/covid19-projections-using-GRM/covid19-projections-using-GRM.Rmd
---

```{r}
# Load the necessary libraries
library(tidyr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(reshape2)
library(gridExtra)
library(deSolve)
library(minpack.lm)
library(incidence)
```


# The data

Will be using data made available by Johns Hopkins Center for Systems Science and Engineering (CSSE)
https://github.com/CSSEGISandData/COVID-19.git


**Confirmed cases**
```{r}
cases <- read.delim('assets/time_series_covid19_confirmed_global.csv', 
                           sep = ',', header = TRUE, stringsAsFactors = FALSE) %>%
  rename(country = Country.Region, province = Province.State, lat = Lat, lon = Long) 
cases <- reshape2::melt(cases, id.vars = c('country', 'province', 'lat', 'lon'),
                             variable.name = 'date', value.name = 'cum_cases') %>%
  mutate(date = as.Date(date, "X%m.%d.%y")) %>% 
  group_by(country, province) %>% 
  mutate(cases = c(cum_cases[1], diff(cum_cases))) %>%
  filter(cases >= 0)
cases$country[cases$country == "Taiwan*"] <- "Taiwan"
cases
```


```{r}
# Confirmed deaths in the world
deaths <- read.delim('assets/time_series_covid19_deaths_global.csv', 
                           sep = ',', header = TRUE, stringsAsFactors = FALSE) %>%
  rename(country = Country.Region, province = Province.State, lat = Lat, lon = Long) 
deaths <- reshape2::melt(deaths, id.vars = c('country', 'province', 'lat', 'lon'),
                             variable.name = 'date', value.name = 'cum_deaths') %>%
  mutate(date = as.Date(date, "X%m.%d.%y")) %>% 
  group_by(country, province) %>% 
  mutate(deaths = c(cum_deaths[1], diff(cum_deaths))) %>%
  filter(deaths >= 0)
deaths$country[deaths$country == "Taiwan*"] <- "Taiwan"
deaths
```

# China

```{r, fig.height=14, fig.width=14}
cases %>% filter(country == "China") %>% 
  ggplot(., aes(x = date, y = cases, color = province)) + 
  geom_line() + 
  facet_wrap(~ province, scales = "free_y") + 
  labs(x = "") + 
  theme_minimal() + 
  theme(legend.position = 'none', 
        axis.text.x = element_text(angle = 90))
```

**Function to fit a Generalized Richards Model to cumulative incidence data**

Define the GRM differential equation given by:
\[dC/dt = r[C(t)]^p(1-C(t)/K)^\alpha\], where
  _C(t)_ = cumulative number of cases at time _t_
  _r_ = growth rate at the early stage
  _p_ \(\epsilon\) [0, 1] = parameter that allows the model to capture different growth profiles including the constant incidence ( _p_ = 0), sub-exponential-growth (0 < _p_ < 1) and exponential growth ( _p_ = 1)
  \(\alpha\) = measures the deviation from the symmetric s-shapes dynamics of the simple logistic curve

The model recovers the original Richards model for _p_ = 1 and reduces to teh generalized logistic model for \(\alpha\) = and _p_ = 1

Source: <a href=https://arxiv.org/ftp/arxiv/papers/2003/2003.05681.pdf>Wu et al. (2020)</a>

```{r}
# Generalized Richards model (GRM)
GRM <- function(time, state, parameters) {
  par <- as.list(c(state, parameters))
  with(par, {
    dC <- r * (C^p) * (1 - (C/K)^a)
    return(list(c(dC)))
  })
}
```

We need a solver to solve the ordinary differential equation and an optimizer for the parameters _K, r, p, \(\alpha\)_. We'll use the function `ode` from the `deSolve` package and the base function `optim` to achieve this. 

```{r}
optimizer <- function(cum_cases, K_start = 10000, r_start = 1, p_start = 0.5, a_start = 1.5) {
  # Solver for the GRM 
  
  RSS <- function(parameters) {
    names(parameters) <- c('K', 'r', 'p', 'a')
    times <- 1:length(cum_cases)
    out <- ode(y = init, times = times, func = GRM, parms = parameters)
    fit <- out[, 2]
    sum((cum_cases - fit)^2)
  }
  
  # Initialize the parameters
  init <- c(C = 1)
  # K_start <- if (is.null(K_start)) median(cum_cases)
  pars <- c(K = K_start, r = r_start, p = p_start, a = a_start)
  
  
  # Optimizer of the parameters using RSS 
  Opt <- optim(pars, RSS, method = "L-BFGS-B", 
               lower = c(0, 0, 0, 0), 
               upper = c(10000000, 10, 1, 20))
  return(Opt)
  }
```

Let's write a function to make predictions based on the fit. As a default, we'll predict cumulative cases 14 days ahead of the latest date data were collected. We'll also use those data and uncumulate them to get incidence predictions.

```{r}
# Prediction function, default is 14 days forward of the last date of observed data
make_preds <- function(df, fit, t_fwd = 14) {
  t <- 1:(as.integer(today() - df$date[1]) + t_fwd)
  # Make predictions based on cumulative incidence
  preds <- data.frame(ode(y = c(C=1), times = t, func = GRM, parms = fit$par))
  preds <- preds %>% 
    mutate(date = df$date[1] + days(t-1),
           C = round(C, 0), 
           # uncumulate to get incidence
           I = c(C[1], diff(C))) %>% 
    left_join(df %>% ungroup() %>% select(date, cum_cases, cases)) 
  return(preds)
}
```

Lastly, create a plotting fucntion. We'll plot both the cumulative and incidence fits against the observed data. 

```{r}
plot_grm_fits <- function(df, name) {
  
  # Cumulative plot of observed vs fitted
  cum_plot <- ggplot(df, aes(x=date)) +
    geom_line(aes(y = C), color = "turquoise", size = 2) +
    geom_point(aes(y = cum_cases), color = "magenta", shape = 1, size = 3) +
    geom_line(aes(y = cum_cases), color = "gray10", linetype = "dashed") + 
    labs(x = "", y = 'Cumulative Incidence', 
         title = paste('Cumulative Cases fitted vs observed in', name),
         subtitle = '(turquoise = fitted incidence, magenta = observed incidence)') +
    theme_minimal()
  
  # Incidence plot of observed vs fitted
  inc_plot <- ggplot(df, aes(x=date)) +
    geom_line(aes(y = I), color = "turquoise", size = 2) +
    geom_point(aes(y = cases), color = "magenta", shape = 1, size = 3) +
    geom_line(aes(y = cases), color = "gray10", linetype = "dashed") +
    labs(y = 'Daily Incidence', title = paste('Incident Cases fitted vs observed in', name),
         subtitle = '(turquoise = fitted incidence, magenta = observed incidence)') +
    theme_minimal()
  
  gridExtra::grid.arrange(cum_plot, inc_plot, ncol = 2)
}
```


Ok, we are ready to get into the fun stuff. Let's start with Hubei since that's the literal epicenter of the pandemic and has way more cases than any other province in China. 

## Hubei

#### Confirmed cases

```{r}
hubei_cases <- cases %>% filter(province == 'Hubei')
plot(hubei_cases$date, hubei_cases$cum_cases, 'l')
```



```{r}
hubei_grm_fit <- optimizer(hubei_cases$cum_cases)
hubei_grm_fit
```
The model converged! Let's make some predictions

```{r, warning=FALSE, message=FALSE, fig.width=10}
hubei_preds <- make_preds(hubei_cases, hubei_grm_fit)
plot_grm_fits(hubei_preds, "Hubei")
```

#### Confirmed deaths

Let's do the same for deaths

```{r}
hubei_deaths <- deaths %>% filter(province == 'Hubei')
plot(hubei_deaths$date, hubei_deaths$cum_deaths, 'l')
```
```{r}
hubei_deaths_grm_fit <- optimizer(hubei_deaths$cum_deaths, K_start = 5000)
hubei_deaths_grm_fit
```
Predictions

```{r}
# Prediction function, default is 14 days forward of the last date of observed data
make_deaths_preds <- function(df, fit, t_fwd = 14) {
  t <- 1:(as.integer(today() - df$date[1]) + t_fwd)
  # Make predictions based on cumulative incidence
  preds <- data.frame(ode(y = c(C=1), times = t, func = GRM, parms = fit$par))
  preds <- preds %>% 
    mutate(date = df$date[1] + days(t-1),
           C = round(C, 0), 
           # uncumulate to get incidence
           I = c(C[1], diff(C))) %>% 
    left_join(df %>% ungroup %>% select(date, cum_deaths, deaths)) 
  return(preds)
}
```

Lastly, create a plotting fucntion. We'll plot both the cumulative and incidence fits against the observed data. 

```{r}
plot_grm_deaths_fits <- function(df, name) {
  
  # Cumulative plot of observed vs fitted
  cum_plot <- ggplot(df, aes(x=date)) +
    geom_line(aes(y = C), color = "red", size = 2) +
    geom_point(aes(y = cum_deaths), color = "orange", shape = 1, size = 3) +
    geom_line(aes(y = cum_deaths), color = "gray10", linetype = "dashed") + 
    labs(x = "", y = 'Cumulative Deaths', 
         title = paste('Cumulative Deaths fitted vs observed in', name),
         subtitle = '(red = fitted incidence, orange = observed incidence)') +
    theme_minimal()
  
  # Incidence plot of observed vs fitted
  inc_plot <- ggplot(df, aes(x=date)) +
    geom_line(aes(y = I), color = "red", size = 2) +
    geom_point(aes(y = deaths), color = "orange", shape = 1, size = 3) +
    geom_line(aes(y = deaths), color = "gray10", linetype = "dashed") +
    labs(y = 'Daily Deaths', title = paste('Incident Deaths fitted vs observed in', name),
         subtitle = '(red = fitted incidence, orange = observed incidence)') +
    theme_minimal()
  
  gridExtra::grid.arrange(cum_plot, inc_plot, ncol = 2)
}
```


```{r, warning=FALSE, message=FALSE, fig.width=10}
hubei_deaths_preds <- make_deaths_preds(hubei_deaths, hubei_deaths_grm_fit)
plot_grm_deaths_fits(hubei_deaths_preds, "Hubei")
```

What's the lag time in the peaks between confirmed cases and deaths

```{r}
hubei_preds %>% filter(I == max(I)) %>% pull(date); 
hubei_deaths_preds %>% filter(I == max(I)) %>% pull(date)
```


## Hong Kong
```{r}
hongkong <- cases %>% filter(province == 'Hong Kong')
plot(hongkong$date, hongkong$cum_cases, 'l')
```
For Hong Kong, we need to provide the starting r to help the model out. Let's use the log-linear fit from the incidence package

```{r}
i.hk <- incidence((hongkong %>% select(date, cases) %>% uncount(cases))$date)
i.hk.peak <- find_peak(i.hk)
i.hk.fit <- incidence::fit(subset(i.hk, from = ymd('2020-02-28')), split = i.hk.peak)
plot(i.hk) %>% add_incidence_fit(i.hk.fit)
i.hk.fit
```

```{r}
hongkong_grm_fit <- optimizer(hongkong$cum_cases, K_start = 200, r_start = 0.14)
hongkong_grm_fit
```

```{r, , fig.width=10, warning=FALSE, message=FALSE}
hongkong_preds <- make_preds(hongkong, hongkong_grm_fit)
plot_grm_fits(hongkong_preds, "Hong Kong")
```
Great! That works. 

## EVerything but Hubei, Hong Kong, Inner Mongolia, Beijing, Shanghai and Heilongjiang
Some of the provinces seem to be getting a second wave of cases. Let's remove these and see if we can use the model 

```{r}
to_rm <- c("Hubei", "Hong Kong", "Beijing", "Shanghai", "Inner Mongolia", "Heilongjiang")
select_china <- cases %>% filter(country == "China", !(province %in% to_rm)) %>% 
  group_by(date) %>% summarise(cum_cases = sum(cum_cases), cases = sum(cases))

plot(select_china$date, select_china$cum_cases)
```

```{r}
i.select_china <- incidence((select_china %>% select(date, cases) %>% uncount(cases))$date)
i.select_china.peak <- find_peak(i.select_china)
i.select_china.fit <- incidence::fit(i.select_china, split = i.select_china.peak)
plot(i.select_china) %>% add_incidence_fit(i.select_china.fit)
i.select_china.fit
```

```{r}
select_china_grm_fit <- optimizer(select_china$cum_cases, K_start = 5000)
select_china_grm_fit
```

```{r, , fig.width=10, warning=FALSE, message=FALSE}
select_china_preds <- make_preds(select_china, select_china_grm_fit)
plot_grm_fits(select_china_preds, paste("China minus\n", paste(to_rm[1:4], collapse = ", ")))
```

# South Korea

```{r}
south_korea <- cases %>% filter(country == "Korea, South") 
plot(south_korea$cum_cases)
```

```{r}
south_korea_grm_fit <- optimizer(south_korea$cum_cases)
south_korea_grm_fit
```

```{r, , fig.width=10, warning=FALSE, message=FALSE}
south_korea_preds <- make_preds(south_korea, south_korea_grm_fit)
plot_grm_fits(south_korea_preds, "South Korea")
```

```{r}
south_korea_deaths <- deaths %>% filter(country == "Korea, South")
plot(south_korea_deaths$cum_deaths)
```


```{r}
south_korea_deaths_grm_fit <- optimizer(south_korea_deaths$cum_deaths, K_start = 10, r_start = 0.10)
south_korea_deaths_grm_fit
```

```{r, warning=FALSE, message=FALSE, fig.width=10}
south_korea_deaths_preds <- make_deaths_preds(south_korea_deaths, south_korea_deaths_grm_fit)
plot_grm_deaths_fits(south_korea_deaths_preds, "South Korea")
```

What's the lag time in the peaks between confirmed cases and deaths

```{r}
ii <- south_korea_preds %>% filter(I == max(I)) %>% pull(date)
dd <- south_korea_deaths_preds %>% filter(I == max(I)) %>% pull(date)
ii[ceiling(length(ii)/2)]; dd[ceiling(length(dd)/2)]
```


# Italy

```{r}
italy <- cases %>% filter(country == "Italy") 
plot(italy$cum_cases)
```

```{r}
italy_grm_fit <- optimizer(italy$cum_cases)
italy_grm_fit
```

```{r, , fig.width=10, warning=FALSE, message=FALSE}
italy_preds <- make_preds(italy, italy_grm_fit)
plot_grm_fits(italy_preds, "Italy")
```

```{r}
italy_deaths <- deaths %>% filter(country == "Italy")
plot(italy_deaths$cum_deaths)
```

The GRM model struggles to fit the death data. We can use the Generalized Logistic Regression model using the `nlsLM` function from the `minpack.lm` package, which is: 
\[C(t) = p/(1+exp(- \alpha\ (t- \beta\ ))\]

Reference: IHME (https://ihmeuw-msca.github.io/CurveFit/methods/)

```{r, message=FALSE, warning=FALSE, fig.width=10}
italy_deaths <- italy_deaths %>% mutate(time = 1:nrow(italy_deaths))

# Fit using the generalized logistic model
italy_deaths_fit <-nlsLM(cum_deaths ~ p/(1 + exp(-a * (time-B))), italy_deaths,
                         start = list(a = 0.2, p = 0.5, B = 10), lower = c(0, 0, 0))

# Make predictions
t <- 1:(as.integer(today() - italy_deaths$date[1]) + 14)
italy_deaths_preds <- data.frame(t = t, C = predict(italy_deaths_fit, list(time=t)))
italy_deaths_preds <- italy_deaths_preds %>%
  mutate(date = italy_deaths$date[1] + days(t-1),
         C = round(C, 0),
         I = c(C[1], diff(C))) %>%
  left_join(italy_deaths %>% ungroup() %>% select(date, cum_deaths, deaths))

# Plot
plot_grm_deaths_fits(italy_deaths_preds, 'Italy')

```



# Spain

```{r}
spain <- cases %>% filter(country == "Spain") 
plot(spain$cum_cases)
```

```{r}
spain_grm_fit <- optimizer(spain$cum_cases)
spain_grm_fit
```

```{r, fig.width=10, warning=FALSE, message=FALSE}
spain_preds <- make_preds(spain, spain_grm_fit)
plot_grm_fits(spain_preds, "Spain")
```

```{r}
spain_deaths <- deaths %>% filter(country == "Spain")
plot(spain_deaths$cum_deaths)
```

```{r, message=FALSE, warning=FALSE, fig.width=10, preview=TRUE}
spain_deaths <- spain_deaths %>% mutate(time = 1:nrow(spain_deaths))

# Fit using the generalized logistic model
spain_deaths_fit <-nlsLM(cum_deaths ~ p/(1 + exp(-a * (time-B))), spain_deaths,
                         start = list(a = 0.2, p = 0.5, B = 10), lower = c(0, 0, 0))

# Make predictions
t <- 1:(as.integer(today() - spain_deaths$date[1]) + 14)
spain_deaths_preds <- data.frame(t = t, C = predict(spain_deaths_fit, list(time=t)))
spain_deaths_preds <- italy_deaths_preds %>%
  mutate(date = italy_deaths$date[1] + days(t-1),
         C = round(C, 0),
         I = c(C[1], diff(C))) %>%
  left_join(spain_deaths %>% ungroup() %>% select(date, cum_deaths, deaths))

# Plot
plot_grm_deaths_fits(spain_deaths_preds, 'Spain')

```

# Germany

```{r}
germany <- cases %>% filter(country == "Germany") 
plot(germany$cum_cases)
```

```{r}
germany_grm_fit <- optimizer(germany$cum_cases)
germany_grm_fit
```

```{r, fig.width=10, warning=FALSE, message=FALSE}
germany_preds <- make_preds(germany, germany_grm_fit)


# Fit using the generalized logistic model
germany_glm_fit <-nlsLM(cum_cases ~ p/(1 + exp(-a * (time-B))), germany_preds,
                         start = list(a = 0.2, p = 0.5, B = 10), lower = c(0, 0, 0))

# Make predictions
germany_preds$glm <- predict(germany_glm_fit, list(time=germany_preds$time))

# Plot
plot_grm_fits(germany_preds, 'Germany') 
ggplot(germany_preds, aes(x = date)) + geom_point(aes(y = cum_cases)) + geom_line(aes(y = glm)) + theme_minimal()

```

In the test run above, the NLS LM using the **Levenberg-Marquardt** is better and faster. 

```{r}
germany_deaths <- deaths %>% filter(country == "Germany")
plot(germany_deaths$cum_deaths)
```

```{r, message=FALSE, warning=FALSE, fig.width=10}
germany_deaths <- germany_deaths %>% mutate(time = 1:nrow(germany_deaths))

# Fit using the generalized logistic model
germany_deaths_fit <-nlsLM(cum_deaths ~ p/(1 + exp(-a * (time-B))), germany_deaths,
                         start = list(a = 0.2, p = 0.5, B = 10), lower = c(0, 0, 0))

# Make predictions
t <- 1:(as.integer(today() - germany_deaths$date[1]) + 14)
germany_deaths_preds <- data.frame(t = t, C = predict(germany_deaths_fit, list(time=t)))
germany_deaths_preds <- germany_deaths_preds %>%
  mutate(date = germany_deaths$date[1] + days(t-1),
         C = round(C, 0),
         I = c(C[1], diff(C))) %>%
  left_join(germany_deaths %>% ungroup() %>% select(date, cum_deaths, deaths))

# Plot
plot_grm_deaths_fits(germany_deaths_preds, 'Germany')

```

Final thoughts: `nlslm` is way faster than using the optim function for GRM and may even be more accurate. Moving forward that's what I'll use. 



